RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-09-03 doc/bioinformatics2013/RVD2.tex
Paper draft update on Moments derivation

\subsection{Initialization}

The initial values of the model parameters and latent variables is obtained by a method-of-moments procedure. Essentially, we set the parameter value (population moment) equal to the sample moment. In some cases, the method-of-moments estimate is outside the parameter space. Then, we XXX.

The estimated error probability is calculated from the observed sample
\begin{equation}
 \theta_{ij}=\frac{r_{ij}}{n_{ij}}
\end{equation}

The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
\begin{equation}
 \mu_j=E(\theta_{ij}|\mu_j,M_j)
\end{equation}
\begin{equation}
 M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
\end{equation}

@@

>>>>

\subsection{Initialization}
The initial values of the model parameters and latent variables is obtained by a method-of-moments procedure. Essentially, we set the parameter value (population moment) equal to the sample moment. In some cases, the method-of-moments estimate is outside the parameter space. Then, we XXX.

The estimated error probability is calculated from the observed sample
\begin{equation}
 \theta_{ij}=\frac{r_{ij}}{n_{ij}}
\end{equation}
The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
\begin{equation}
 \mu_j=E(\theta_{ij}|\mu_j,M_j)
\end{equation}
\begin{equation}
 M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
\end{equation}

@@

\section{Appendix}
\subsection{Moments derivation}
The probability density function of beta distribution, for $x\in[0,1]$, and shape real parameter $\alpha>1$ and $\beta>1$ is 
\begin{equation}
\begin{split}
  f(x;\alpha,\beta) &= \frac{x^{\alpha-1}(1-x)^{\beta-1}}{\int_0^1x^{\alpha-1}(1-x)^{\beta-1}dx}\\ &=\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}x^{\alpha-1}(1-x)^{\beta-1}\\ &=\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}
\end{split}
\end{equation}
where $\Gamma(z)$ is the Gammma function, $B(\alpha,\beta)$ is the beta function, and
\begin{equation}
\begin{split}
 B(\alpha,\beta) &=\int_0^1x^{\alpha-1}(1-x)^{\beta-1}dx\\  &= \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha+\beta)}
\end{split}
\end{equation}

The expected value (mean) of a Beta distribution random variable X with parameter $\alpha$ and $\beta$ is 
\begin{equation}
\begin{split}
  E[X] &= \int_0^1xf(x;\alpha,\beta)dx\\ &= \int_0^1x\frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}dx\\ &= \frac{\int_0^1x^{\alpha}(1-x)^{\beta-1}dx}{B(\alpha,\beta)}\\ &= \frac{B(\alpha+1,\beta)}{B(\alpha,\beta)}\\ &= \frac{\Gamma(\alpha+1)\Gamma(\beta)}{\Gamma(\alpha+\beta+1)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\\ &= \frac{\alpha\Gamma(\alpha)\Gamma(\beta)}{(\alpha+\beta)\Gamma(\alpha+\beta)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\\ &= \frac{\alpha}{\alpha+\beta}
\end{split}
\end{equation}
Note that Gamma function satisfies the functional equation
\begin{equation}
 \Gamma(z) = z\Gamma(z)
\end{equation}

The variance of a Beta distribution random variable X with parameter $\alpha$ and $\beta$ is 
\begin{equation}
\begin{split}
  var(X) &= E[(X-E(X))^2]\\ &= E(X^2)-[E(x)]^2\\ &=\int_0^1x^2\frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}dx-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\int_0^1x^{\alpha+1}(1-x)^{\beta-1}dx}{B(\alpha,\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{B(\alpha+2,\beta)}{B(\alpha,\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\Gamma(\alpha+2)\Gamma(\beta)}{\Gamma(\alpha+\beta+2)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{(\alpha+1)\alpha\Gamma(\alpha)\Gamma(\beta)}{(\alpha+\beta+1)(\alpha+\beta)\Gamma(\alpha+\beta)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{(\alpha+1)\alpha}{(\alpha+\beta+1)(\alpha+\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\alpha\beta}{(\alpha+\beta+1)(\alpha+\beta)^2}
\end{split}
\end{equation}

Instead of the standard scale parameter $\alpha$ and $\beta$, we use the mean $\mu$ and sample-size paramenter $M$ for the Beta distribution in our hierarchical model.
\begin{equation}
\begin{split}
  f(x;\mu,M) &=\frac{\Gamma(M)}{\Gamma(M\mu)\Gamma(M(1-\mu))}x^{M\mu-1}(1-x)^{M(1-\mu)-1}\\ &= \frac{1}{B(M\mu,M(1-\mu))}x^{M\mu-1}(1-x)^{M(1-\mu)-1}
\end{split}
\end{equation}
where 
\begin{equation}
 \mu = \frac{\alpha}{\alpha+\beta}
\end{equation}
\begin{equation}
 M = \alpha+\beta
\end{equation}
From Equation 24 and Equation 26 it can be seen that
\begin{equation}
 E(x|\mu,M) = \mu
\end{equation}
\begin{equation}
 Var(x|\mu,M) = \frac{\mu(1-\mu)}{M+1}
\end{equation}
